Loop-Aware Optimizations in PyPy's Tracing JIT

### 2012-08-16 talk/dls2012/paper.tex
update the correct paper
make figure wide
this is done
kill the mention of psyco and analyze luajit a tiny bit

\revb{
A nit: Section 7 says that loop peeling never makes runtime
performance worse, but generating more code can potentially slow
performance. I assume that non-numeric benchmarks show no slowdown in
practice, and that might be worth noting.
}

@@

\begin{figure}
\begin{center}
{\smaller
\begin{tabular}{|l|r|r|r|r|r|r|r|}
\hline
 & CPython & PyPy  & PyPy & GCC & luajit & luajit \\
 &        & no LP &       & -O3 &        & no LP \\
FFT(1024, 32768) & 469.07 & 20.83 +- 0.02 & 12.73 +- 0.01 & - & 2.76 +- 0.01 & 4.42 +- 0.01\\
\hline
FFT(1024,32768) & - & - & - & 1.40 +- 0.04 & - & -\\
\hline
FFT(1048576, 2) & 58.93 & 4.12 +- 0.01 & 2.05 +- 0.00 & - & 1.08 +- 0.01 & 1.33 +- 0.01\\
\hline
FFT(1048576,2) & - & - & - & 0.83 +- 0.02 & - & -\\
\hline
LU(100, 4096) & 1974.14 & 32.22 +- 0.14 & 13.39 +- 0.03 & - & 1.52 +- 0.01 & 8.57 +- 0.01\\
\hline
LU(100,4096) & - & - & - & 1.33 +- 0.04 & - & -\\
\hline
LU(1000, 2) & 955.31 & 14.98 +- 0.22 & 5.99 +- 0.21 & - & 0.67 +- 0.01 & 3.99 +- 0.01\\
\hline
LU(1000,2) & - & - & - & 0.65 +- 0.04 & - & -\\
\hline
MonteCarlo(268435456) & 618.89 & 20.60 +- 0.05 & 15.33 +- 0.08 & 1.69 +- 0.05 & 2.82 +- 0.00 & 3.92 +- 0.01\\
\hline
SOR(100, 32768) & 1458.12 & 8.24 +- 0.00 & 2.66 +- 0.00 & - & 1.31 +- 0.01 & 2.02 +- 0.00\\
\hline
SOR(100,32768) & - & - & - & 1.76 +- 0.04 & - & -\\
\hline
SOR(1000, 256) & 1210.45 & 6.48 +- 0.00 & 2.10 +- 0.00 & - & 1.08 +- 0.01 & 1.63 +- 0.00\\
\hline
SOR(1000,256) & - & - & - & 1.49 +- 0.02 & - & -\\
\hline
SparseMatMult(1000, 5000, 262144) & 371.66 & 24.25 +- 0.04 & 16.52 +- 0.04 & - & - & -\\
\hline
SparseMatMult(1000,5000,262144) & - & - & - & 1.84 +- 0.03 & 4.53 +- 0.02 & 9.64 +- 0.02\\
\hline
SparseMatMult(100000, 1000000, 1024) & 236.93 & 17.01 +- 0.01 & 8.75 +- 0.08 & - & - & -\\
\hline
SparseMatMult(100000,1000000,1024) & - & - & - & 1.20 +- 0.03 & 2.42 +- 0.01 & 7.19 +- 0.01\\
\hline
conv3(100, nil) & - & - & - & - & 0.18 +- 0.00 & 0.70 +- 0.00\\
\hline
conv3(1000, nil) & - & - & - & - & 0.12 +- 0.01 & 0.67 +- 0.01\\
\hline
conv3(1e5) & - & - & - & 0.52 +- 0.04 & - & -\\
\hline
conv3(1e6) & - & - & - & 0.60 +- 0.03 & - & -\\
\hline
conv3(array(1e5)) & 50.14 & 1.09 +- 0.01 & 0.49 +- 0.01 & - & - & -\\
\hline
conv3(array(1e6)) & 49.20 & 1.13 +- 0.02 & 0.51 +- 0.00 & - & - & -\\
\hline
conv3x3(1000) & - & - & - & 0.17 +- 0.04 & - & -\\
\hline
conv3x3(1000, 1000) & - & - & - & - & 0.09 +- 0.01 & 0.49 +- 0.01\\
\hline
conv3x3(1000000, 3) & - & - & - & - & 0.13 +- 0.00 & 0.53 +- 0.00\\
\hline
conv3x3(3) & - & - & - & 0.19 +- 0.03 & - & -\\
\hline
conv3x3(Array2D(1000000x3)) & 139.81 & 0.70 +- 0.00 & 0.21 +- 0.00 & - & - & -\\
\hline
conv3x3(Array2D(1000x1000)) & 138.95 & 0.70 +- 0.00 & 0.20 +- 0.00 & - & - & -\\
\hline
conv5(100, nil) & - & - & - & - & 0.21 +- 0.01 & 0.87 +- 0.01\\
\hline
conv5(1000, nil) & - & - & - & - & 0.17 +- 0.01 & 0.84 +- 0.00\\
\hline
conv5(1e5) & - & - & - & 0.55 +- 0.02 & - & -\\
\hline
conv5(1e6) & - & - & - & 0.58 +- 0.03 & - & -\\
\hline
conv5(array(1e5)) & 74.65 & 1.22 +- 0.00 & 0.64 +- 0.00 & - & - & -\\
\hline
conv5(array(1e6)) & 77.94 & 1.26 +- 0.00 & 0.68 +- 0.01 & - & - & -\\
\hline
dilate3x3(1000) & - & - & - & 0.17 +- 0.03 & - & -\\
\hline
dilate3x3(1000, 1000) & - & - & - & - & 0.09 +- 0.00 & 0.48 +- 0.01\\
\hline
dilate3x3(Array2D(1000x1000)) & 137.52 & 4.35 +- 0.01 & 3.91 +- 0.02 & - & - & -\\
\hline
sobel(Array2D(1000x1000)) & 104.02 & 0.49 +- 0.00 & 0.21 +- 0.00 & 0.17 +- 0.03 & - & -\\
\hline
sobel\_magnitude(1000, 1000) & - & - & - & - & 0.24 +- 0.01 & 0.60 +- 0.01\\
\hline
sqrt(Fix16) & 463.46 & 5.12 +- 0.00 & 2.96 +- 0.00 & 1.34 +- 0.03 & 1.14 +- 0.00 & 12.80 +- 0.04\\
\hline
sqrt(float) & 14.99 & 1.37 +- 0.00 & 0.89 +- 0.00 & 0.85 +- 0.04 & 0.83 +- 0.01 & 1.06 +- 0.01\\
\hline
sqrt(int) & 13.91 & 3.22 +- 0.02 & 2.65 +- 0.00 & 1.25 +- 0.03 & 0.83 +- 0.01 & 1.06 +- 0.00\\
\hline
\end{tabular}
}
\end{center}
\label{fig:benchmarks}
\caption{Benchmark Results in Seconds. Arrays of length $10^5$ and
  $10^6$ and matrices of size $1000\times 1000$ and $1000000 \times
  3$ are used. The one used in each benchmark is indicated in
  the leftmost column. For the matrices, only the number of rows are
  specified.} 
\end{figure}

@@

We run GCC with -O3 -march=native, disabling the
automatic loop vectorization. In all cases, SSE2 instructions were used for
floating point operations, except Psyco which uses x87 FPU instructions.
% Psyco does not use the x87 FPU: all floating-point arithmetic is done with
% residual calls to C helpers.  These can probably be compiled with SSE2.
% But compiling CPython (and maybe Psyco) for x87 or SSE2 has probably
% no measurable effect.
We also run PyPy with loop peeling optimization and without (but otherwise
identical).

For PyPy and Lua 10 iterations were run, prefaced with 3 iterations for warming up.
Due to benchmarks taking large amounts of time on CPython, only one run
was performed, prefaced with one warmup run for Psyco.
For GCC 5 iterations
were run. In all cases, the standard deviation is very low, making benchmarks
very well reproducible.

We can observe that PyPy (even without loop peeling) is orders of magnitude
faster than either CPython or Psyco. This is due to the JIT compilation
advantages and optimizations we discussed in previous
work~\cite{bolz_allocation_2011, bolz_runtime_2011}. The geometric mean of the
speedup of loop peeling is 70\%, which makes benchmark times
comparable with native-compiled C code. We attribute the performance gap to C code to
the relative immaturity of RPython's JIT machine code backend as well as missing
optimizations, like instruction scheduling. Also, in case of nested loops, 
operations are only moved out of the 
innermost loop. That is an issue when the innermost loop is 
short and a significant amount of time is spent in the outer loops. This is the case 
with for example SparseMatMult.

>>>>

@@

\begin{figure*}
\begin{center}
{\smaller
\begin{tabular}{|l|r|r|r|r|r|r|r|}
\hline
 & CPython & PyPy  & PyPy & GCC & luajit & luajit \\
 &        & no LP &       & -O3 &        & no LP \\
FFT(1024, 32768) & 469.07 & 20.83 +- 0.02 & 12.73 +- 0.01 & - & 2.76 +- 0.01 & 4.42 +- 0.01\\
\hline
FFT(1024,32768) & - & - & - & 1.40 +- 0.04 & - & -\\
\hline
FFT(1048576, 2) & 58.93 & 4.12 +- 0.01 & 2.05 +- 0.00 & - & 1.08 +- 0.01 & 1.33 +- 0.01\\
\hline
FFT(1048576,2) & - & - & - & 0.83 +- 0.02 & - & -\\
\hline
LU(100, 4096) & 1974.14 & 32.22 +- 0.14 & 13.39 +- 0.03 & - & 1.52 +- 0.01 & 8.57 +- 0.01\\
\hline
LU(100,4096) & - & - & - & 1.33 +- 0.04 & - & -\\
\hline
LU(1000, 2) & 955.31 & 14.98 +- 0.22 & 5.99 +- 0.21 & - & 0.67 +- 0.01 & 3.99 +- 0.01\\
\hline
LU(1000,2) & - & - & - & 0.65 +- 0.04 & - & -\\
\hline
MonteCarlo(268435456) & 618.89 & 20.60 +- 0.05 & 15.33 +- 0.08 & 1.69 +- 0.05 & 2.82 +- 0.00 & 3.92 +- 0.01\\
\hline
SOR(100, 32768) & 1458.12 & 8.24 +- 0.00 & 2.66 +- 0.00 & - & 1.31 +- 0.01 & 2.02 +- 0.00\\
\hline
SOR(100,32768) & - & - & - & 1.76 +- 0.04 & - & -\\
\hline
SOR(1000, 256) & 1210.45 & 6.48 +- 0.00 & 2.10 +- 0.00 & - & 1.08 +- 0.01 & 1.63 +- 0.00\\
\hline
SOR(1000,256) & - & - & - & 1.49 +- 0.02 & - & -\\
\hline
SparseMatMult(1000, 5000, 262144) & 371.66 & 24.25 +- 0.04 & 16.52 +- 0.04 & - & - & -\\
\hline
SparseMatMult(1000,5000,262144) & - & - & - & 1.84 +- 0.03 & 4.53 +- 0.02 & 9.64 +- 0.02\\
\hline
SparseMatMult(100000, 1000000, 1024) & 236.93 & 17.01 +- 0.01 & 8.75 +- 0.08 & - & - & -\\
\hline
SparseMatMult(100000,1000000,1024) & - & - & - & 1.20 +- 0.03 & 2.42 +- 0.01 & 7.19 +- 0.01\\
\hline
conv3(100, nil) & - & - & - & - & 0.18 +- 0.00 & 0.70 +- 0.00\\
\hline
conv3(1000, nil) & - & - & - & - & 0.12 +- 0.01 & 0.67 +- 0.01\\
\hline
conv3(1e5) & - & - & - & 0.52 +- 0.04 & - & -\\
\hline
conv3(1e6) & - & - & - & 0.60 +- 0.03 & - & -\\
\hline
conv3(array(1e5)) & 50.14 & 1.09 +- 0.01 & 0.49 +- 0.01 & - & - & -\\
\hline
conv3(array(1e6)) & 49.20 & 1.13 +- 0.02 & 0.51 +- 0.00 & - & - & -\\
\hline
conv3x3(1000) & - & - & - & 0.17 +- 0.04 & - & -\\
\hline
conv3x3(1000, 1000) & - & - & - & - & 0.09 +- 0.01 & 0.49 +- 0.01\\
\hline
conv3x3(1000000, 3) & - & - & - & - & 0.13 +- 0.00 & 0.53 +- 0.00\\
\hline
conv3x3(3) & - & - & - & 0.19 +- 0.03 & - & -\\
\hline
conv3x3(Array2D(1000000x3)) & 139.81 & 0.70 +- 0.00 & 0.21 +- 0.00 & - & - & -\\
\hline
conv3x3(Array2D(1000x1000)) & 138.95 & 0.70 +- 0.00 & 0.20 +- 0.00 & - & - & -\\
\hline
conv5(100, nil) & - & - & - & - & 0.21 +- 0.01 & 0.87 +- 0.01\\
\hline
conv5(1000, nil) & - & - & - & - & 0.17 +- 0.01 & 0.84 +- 0.00\\
\hline
conv5(1e5) & - & - & - & 0.55 +- 0.02 & - & -\\
\hline
conv5(1e6) & - & - & - & 0.58 +- 0.03 & - & -\\
\hline
conv5(array(1e5)) & 74.65 & 1.22 +- 0.00 & 0.64 +- 0.00 & - & - & -\\
\hline
conv5(array(1e6)) & 77.94 & 1.26 +- 0.00 & 0.68 +- 0.01 & - & - & -\\
\hline
dilate3x3(1000) & - & - & - & 0.17 +- 0.03 & - & -\\
\hline
dilate3x3(1000, 1000) & - & - & - & - & 0.09 +- 0.00 & 0.48 +- 0.01\\
\hline
dilate3x3(Array2D(1000x1000)) & 137.52 & 4.35 +- 0.01 & 3.91 +- 0.02 & - & - & -\\
\hline
sobel(Array2D(1000x1000)) & 104.02 & 0.49 +- 0.00 & 0.21 +- 0.00 & 0.17 +- 0.03 & - & -\\
\hline
sobel\_magnitude(1000, 1000) & - & - & - & - & 0.24 +- 0.01 & 0.60 +- 0.01\\
\hline
sqrt(Fix16) & 463.46 & 5.12 +- 0.00 & 2.96 +- 0.00 & 1.34 +- 0.03 & 1.14 +- 0.00 & 12.80 +- 0.04\\
\hline
sqrt(float) & 14.99 & 1.37 +- 0.00 & 0.89 +- 0.00 & 0.85 +- 0.04 & 0.83 +- 0.01 & 1.06 +- 0.01\\
\hline
sqrt(int) & 13.91 & 3.22 +- 0.02 & 2.65 +- 0.00 & 1.25 +- 0.03 & 0.83 +- 0.01 & 1.06 +- 0.00\\
\hline
\end{tabular}
}
\end{center}
\label{fig:benchmarks}
\caption{Benchmark Results in Seconds. Arrays of length $10^5$ and
  $10^6$ and matrices of size $1000\times 1000$ and $1000000 \times
  3$ are used. The one used in each benchmark is indicated in
  the leftmost column. For the matrices, only the number of rows are
  specified.} 
\end{figure*}

@@

We run GCC with -O3 -march=native, disabling the
automatic loop vectorization. In all cases, SSE2 instructions were used for
floating point operations.
We also run PyPy and LuaJIT with loop peeling optimization and without (but otherwise
identical).

For PyPy and LuaJIT 10 iterations were run, prefaced with 3 iterations for warming up.
Due to benchmarks taking large amounts of time on CPython, only one run
was performed.
For GCC 5 iterations
were run. In all cases, the standard deviation is very low, making benchmarks
very well reproducible.

We can observe that PyPy (even without loop peeling) is orders of magnitude
faster than CPython. This is due to the JIT compilation
advantages and optimizations we discussed in previous
work~\cite{bolz_allocation_2011, bolz_runtime_2011}. The geometric mean of the
speedup of loop peeling is 70\%, which makes benchmark times
comparable with native-compiled C code. We attribute the performance gap to C code to
the relative immaturity of RPython's JIT machine code backend as well as missing
optimizations, like instruction scheduling. Also, in case of nested loops, 
operations are only moved out of the 
innermost loop. That is an issue when the innermost loop is 
short and a significant amount of time is spent in the outer loops. This is the case 
with for example SparseMatMult.

The speedups that LuaJIT gains from the loop optimization pass are similar to
those PyPy gains. In general, LuaJIT is even closer to C performance, sometimes
even surpassing it. LuaJIT is generating machine code of higher quality because
it has a much better register allocator than PyPy, among other things.
